Semantic Linearity in a Lazy Optimising Compiler

### 2025-01-29 icfp.tex
comments on the proof
added well-typed state and env expansion

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
    \qquad
    \infer[]
    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
    e'[\pi/p]\Downarrow \Theta : v}
    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
    \Delta : e'[x/y] \Downarrow \Theta : v }
    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : v}
    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : K~\ov{x_i} \Downarrow \Gamma : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
    \Delta : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
    \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
    \Delta : e'[K~\ov{x_i}/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
    : v
    }
    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta : v}
    \end{array}
\]

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a linearity
aware operational semantics in which linear variables are erased from
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is type-aware, defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$.

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
    \mid v) : \sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau}\\[1.5em]

       \infer[]
    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
    (\Theta \mid K~\ov{x_i}) : T}
    \\[1.5em]

    \infer[]
    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
    K~\ov{x_i}) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

@@

 \begin{definition}[Well-typed State]
   We write $\vdash \Theta \mid e : \tau$ to denote:

   \[
\vdash \llet{\Theta}{e} : \tau
   \]

 \end{definition}

\begin{theorem}[Type Preservation]
  Let $(\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau$.
  If $\vdash \Theta \mid e : \tau$ then $\vdash \Theta' \mid v : \tau$. 
 \end{theorem}

>>>>

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
    \qquad
    \infer[]
    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
    e'[\pi/p]\Downarrow \Theta : v}
    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
    \Delta : e'[x/y] \Downarrow \Theta : v }
    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : v}
    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : K~\ov{x_i} \Downarrow \Gamma : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
    \Delta : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
    \Delta : e'[K~\ov{x_i}/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
    : v
    }
    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta : v}
    \end{array}
\]

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a linearity
aware operational semantics in which linear variables are erased from
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is type-aware, defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which variables in $\Theta$ are used.

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma, \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
    \mid v) : \sigma , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]

       \infer[]
    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
    \\[1.5em]

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, \ov{x_i :_\omega \sigma_i} , z:_{\Delta'} T ;\Delta , \ov{x_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' :
    K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_i} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma, z:_{\Delta'} T ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e' , \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    % \infer[]
    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % % \Downarrow \Theta_n : v_n \quad
    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    % e_i} \mid e' )\Downarrow (\Theta'
    % \mid v) : \tau
    % }
    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    % \Downarrow (\Theta' \mid v) : \tau}

@@

 We write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.

 \begin{definition}[Environment Expansion]
   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
   \[
     \begin{array}{lcl}
       \left[ \cdot \mid e \right] & \triangleq & e\\
       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & (\lambda x_1:\sigma . [\Theta \mid e])~e'\\
       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & \llet{x:_\Delta \sigma = e'}{[\Theta \mid e]}
    \end{array}
   \]
 \end{definition}

 \begin{definition}[Well-typed State]
   We write $\G;\D \vdash \Theta \mid e : \tau,\Sigma$ to denote the following typing judgment:

   \[
\G ; \D \vdash [ \Theta \mid (e,\mathit{terms}(\Sigma))] : \tau \otimes \bigotimes(\Sigma)
   \]
   where $\mathit{terms}(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ stands for the tuple $(e_1 ,( \dots , (e_n , ())))$ and
   $\bigotimes(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ for $\sigma_1 \otimes ( \dots (\sigma_n \otimes ()))$.
 \end{definition}

 \begin{lemma}
   Let $\G ; \D , \ov{x_i :_{\Delta_i} \tau_i} ; \ov{x_j :_1 : \tau_j} \vdash  e : \sigma$, we have that
   $\G ; \D , \ov{\Delta_i} \vdash [ \ov{x_i :_{\Delta_i} \tau_i = e_i} , \ov{x_j :_1 : \tau_j = e_j} \mid e] : \sigma$
   \end{lemma}

\begin{theorem}[Type Preservation]
  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau , \Sigma$.
  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$

\end{theorem}

% G ; D |- (Theta | e x) \Downarrow (Theta'' | v) : tau, Sigma

% QUEREMOS PROVAR:
% Se
% G ; D |- (Theta | e x) : tau , Sigma
% entao
% G ; D |- (Theta'' | v) : tau, Sigma

% por i.h.
% Se G;D |- (Theta | e) : sigma -> tau, x:tau , Sigma
% entao
% G ; D |- (Theta' | \y:sigma.e') : sigma->tau, x:tau,Sigma

% por i.h.
% Se G ; D |- (Theta' , y:sigma = x | e') : tau, Sigma
% entao
% G;D |- (Theta'' | v) : tau , Sigma

% Gamma ; Delta |- [Theta | (e x , terms(Sigma))] : (tau , types(Sigma))
% Gamma ; Delta |- [Theta | (e , terms(x:tau,Sigma))] : (tau->sigma , types(x:tau,Sigma))

% Daqui Gamma ; Delta |- [Theta' | (\y:_1 sigma.e' , terms(x:tau,Sigma))] : (sigma->tau, types(x:tau,Sigma))
% tirar Gamma ; Delta |- [Theta',y:sigma = x | (e' , terms(Sigma))] : (sigma->tau, types(Sigma))

% [Theta' , x_1 : tau = y | e] = (\x_1:tau . [Theta' | e] ) y
% [ Theta' , x_Delta:tau = e' | e] = let x_Delta = e' in [Theta' | e]
% [ . | (e,e')] = (e,e')

% Gamma ; Delta |- [Theta | (e , terms(Sigma))] : (tau , types(Sigma))

 % Gamma ; Delta |- Theta | e  : tau, Sigma
